---
title: "CytoCrunch: Flow Cytometry Analysis Pipeline"
subtitle: "Clustering, Dimensionality Reduction & Visualization Demo"
author: "Jason Luthy"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    self_contained: true
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: false
    code_folding: show
    df_print: paged
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../reports")
  })
---

```{css, echo=FALSE}
/* ── CytoCrunch Dark Theme ─────────────────────────────────────────────────── */
/* Matches jluthy.github.io portfolio aesthetic                                */

@import url('https://fonts.googleapis.com/css2?family=Space+Mono:ital,wght@0,400;0,700;1,400&family=Syne:wght@400;600;700;800&family=DM+Sans:ital,opsz,wght@0,9..40,300;0,9..40,400;0,9..40,500;1,9..40,300&display=swap');

:root {
  --bg: #050a0e;
  --surface: #0b1520;
  --surface2: #111d2b;
  --border: #1a2e42;
  --accent: #00d4ff;
  --accent2: #7b2fff;
  --accent3: #00ff9d;
  --text: #e8f4f8;
  --text-muted: #6b8fa8;
  --text-dim: #3a5a70;
}

/* ── Base ── */
body {
  background: var(--bg) !important;
  color: var(--text) !important;
  font-family: 'DM Sans', -apple-system, sans-serif !important;
  font-size: 16px;
  line-height: 1.75;
  margin: 0;
  padding: 0;
}

body::before {
  content: '';
  position: fixed;
  inset: 0;
  background-image:
    linear-gradient(rgba(0,212,255,0.025) 1px, transparent 1px),
    linear-gradient(90deg, rgba(0,212,255,0.025) 1px, transparent 1px);
  background-size: 40px 40px;
  pointer-events: none;
  z-index: 0;
}

/* ── Main content container ── */
.main-container {
  max-width: 960px !important;
  margin: 0 auto;
  padding: 2rem 3rem;
  position: relative;
  z-index: 1;
}

/* ── Typography ── */
h1, h2, h3, h4, h5, h6 {
  font-family: 'Syne', sans-serif !important;
  color: var(--text) !important;
  font-weight: 700;
  letter-spacing: -0.02em;
}

h1.title {
  font-size: 2.8rem !important;
  font-weight: 800 !important;
  margin-bottom: 0.25rem !important;
  background: linear-gradient(135deg, var(--accent), var(--accent2));
  -webkit-background-clip: text;
  -webkit-text-fill-color: transparent;
  background-clip: text;
  padding-top: 2rem;
}

h3.subtitle {
  font-family: 'Space Mono', monospace !important;
  font-size: 0.85rem !important;
  font-weight: 400 !important;
  letter-spacing: 0.12em;
  text-transform: uppercase;
  color: var(--text-muted) !important;
  margin-bottom: 0.25rem !important;
}

h4.author {
  font-family: 'Space Mono', monospace !important;
  font-size: 0.75rem !important;
  font-weight: 400 !important;
  letter-spacing: 0.1em;
  color: var(--text-dim) !important;
  margin-bottom: 0.15rem !important;
}

h4.date {
  font-family: 'Space Mono', monospace !important;
  font-size: 0.7rem !important;
  font-weight: 400 !important;
  color: var(--text-dim) !important;
  margin-bottom: 2rem !important;
}

h1 { font-size: 2rem !important; margin-top: 3.5rem !important; margin-bottom: 1.5rem !important; }
h2 { font-size: 1.5rem !important; margin-top: 2.5rem !important; margin-bottom: 1rem !important; color: var(--accent) !important; }
h3 { font-size: 1.15rem !important; margin-top: 2rem !important; margin-bottom: 0.75rem !important; }

/* Section dividers */
h1::before {
  content: '//';
  font-family: 'Space Mono', monospace;
  font-size: 0.8rem;
  color: var(--accent);
  letter-spacing: 0.1em;
  display: block;
  margin-bottom: 0.5rem;
}

p, li {
  color: var(--text-muted) !important;
  font-size: 1rem;
  line-height: 1.75;
}

p strong, li strong {
  color: var(--text) !important;
  font-weight: 500;
}

a { color: var(--accent) !important; text-decoration: none; }
a:hover { text-decoration: underline; }

/* ── TOC sidebar ── */
.tocify-wrapper {
  background: var(--surface) !important;
  border: 1px solid var(--border) !important;
  border-radius: 0 !important;
}

.tocify {
  background: var(--surface) !important;
  border: 1px solid var(--border) !important;
  border-radius: 0 !important;
}

.tocify ul, .tocify li {
  background: var(--surface) !important;
}

.tocify li {
  border: none !important;
}

.tocify a {
  font-family: 'Space Mono', monospace !important;
  font-size: 0.72rem !important;
  letter-spacing: 0.06em;
  color: var(--text-muted) !important;
  padding: 6px 12px !important;
}

.tocify .active {
  background: var(--surface2) !important;
  border-left: 2px solid var(--accent) !important;
}

.tocify .active a {
  color: var(--accent) !important;
}

.tocify-header {
  background: var(--surface) !important;
}

.tocify-subheader {
  background: var(--surface) !important;
}

/* ── Code blocks ── */
pre {
  background: var(--surface) !important;
  color: var(--text) !important;
  border: 1px solid var(--border) !important;
  border-left: 3px solid var(--accent2) !important;
  border-radius: 0 !important;
  padding: 1.25rem !important;
  font-family: 'Space Mono', monospace !important;
  font-size: 0.82rem !important;
  line-height: 1.65 !important;
  overflow-x: auto;
}

code {
  font-family: 'Space Mono', monospace !important;
  font-size: 0.82rem !important;
  color: var(--accent3) !important;
  background: var(--surface) !important;
  padding: 0.15em 0.4em;
  border-radius: 2px;
}

pre code {
  padding: 0 !important;
  background: transparent !important;
  color: var(--text) !important;
}

/* Code fold buttons */
.code-folding-btn,
.btn-default {
  background: var(--surface2) !important;
  color: var(--text-muted) !important;
  border: 1px solid var(--border) !important;
  border-radius: 0 !important;
  font-family: 'Space Mono', monospace !important;
  font-size: 0.7rem !important;
  letter-spacing: 0.06em;
}

.code-folding-btn:hover,
.btn-default:hover {
  background: var(--border) !important;
  color: var(--accent) !important;
}

/* ── Syntax highlighting ── */
pre .kw  { color: var(--accent2) !important; font-weight: 700; }  /* keywords */
pre .fu  { color: var(--accent) !important; }                      /* functions */
pre .st  { color: var(--accent3) !important; }                     /* strings */
pre .dv  { color: #ff922b !important; }                            /* numbers */
pre .co  { color: var(--text-dim) !important; font-style: italic; } /* comments */
pre .ot  { color: var(--accent) !important; }                      /* operators */

/* R output */
pre.r {
  border-left-color: var(--accent2) !important;
}

/* Console output */
pre:not(.r) {
  border-left-color: var(--text-dim) !important;
  color: var(--text-muted) !important;
}

/* ── Tables ── */
table {
  width: 100%;
  border-collapse: collapse;
  margin: 1.5rem 0;
  font-size: 0.9rem;
}

th {
  background: var(--surface2) !important;
  color: var(--accent) !important;
  font-family: 'Space Mono', monospace !important;
  font-size: 0.72rem !important;
  letter-spacing: 0.08em;
  text-transform: uppercase;
  padding: 0.75rem 1rem !important;
  border-bottom: 2px solid var(--accent) !important;
  text-align: left;
}

td {
  padding: 0.6rem 1rem !important;
  border-bottom: 1px solid var(--border) !important;
  color: var(--text-muted) !important;
}

tr:hover td {
  background: var(--surface) !important;
  color: var(--text) !important;
}

/* Paged tables */
.pagedtable-wrapper {
  border: 1px solid var(--border) !important;
  border-radius: 0 !important;
}

.pagedtable {
  background: var(--surface) !important;
}

/* ── Plot output ── */
.figure {
  margin: 2rem 0;
}

img {
  max-width: 100%;
  border: 1px solid var(--border);
  background: var(--surface);
}

p.caption {
  font-family: 'Space Mono', monospace !important;
  font-size: 0.72rem !important;
  color: var(--text-dim) !important;
  letter-spacing: 0.06em;
  margin-top: 0.5rem;
}

/* ── Callout boxes (custom) ── */
.callout-note {
  background: var(--surface);
  border: 1px solid var(--border);
  border-left: 3px solid var(--accent);
  padding: 1.25rem 1.5rem;
  margin: 1.5rem 0;
  font-size: 0.95rem;
}

.callout-note p {
  margin: 0;
}

.callout-note strong {
  color: var(--accent) !important;
  font-family: 'Space Mono', monospace;
  font-size: 0.75rem;
  letter-spacing: 0.08em;
  text-transform: uppercase;
}

/* ── Horizontal rules ── */
hr {
  border: none;
  height: 1px;
  background: var(--border);
  margin: 3rem 0;
}

/* ── Scrollbar ── */
::-webkit-scrollbar { width: 6px; height: 6px; }
::-webkit-scrollbar-track { background: var(--bg); }
::-webkit-scrollbar-thumb { background: var(--border); }
::-webkit-scrollbar-thumb:hover { background: var(--text-dim); }

/* ── Responsive ── */
@media (max-width: 768px) {
  .main-container { padding: 1rem 1.25rem; }
  h1.title { font-size: 2rem !important; }
  pre { font-size: 0.75rem !important; }
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 10,
  fig.height = 6,
  dpi        = 150,
  comment    = "#>"
)

# ── Custom ggplot2 theme matching the site aesthetic ──────────────────────────
library(ggplot2)

theme_cytoCrunch <- function(base_size = 13) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      # Background
      plot.background    = element_rect(fill = "#0b1520", color = NA),
      panel.background   = element_rect(fill = "#0b1520", color = NA),
      panel.grid.major   = element_line(color = "#1a2e42", linewidth = 0.3),
      panel.grid.minor   = element_blank(),

      # Text
      text               = element_text(color = "#e8f4f8", family = "sans"),
      plot.title         = element_text(color = "#e8f4f8", face = "bold", size = rel(1.2),
                                         hjust = 0, margin = margin(b = 8)),
      plot.subtitle      = element_text(color = "#6b8fa8", size = rel(0.85),
                                         hjust = 0, margin = margin(b = 15)),
      axis.text          = element_text(color = "#6b8fa8", size = rel(0.8)),
      axis.title         = element_text(color = "#6b8fa8", size = rel(0.85)),
      legend.text        = element_text(color = "#6b8fa8", size = rel(0.8)),
      legend.title       = element_text(color = "#e8f4f8", size = rel(0.85), face = "bold"),

      # Legend
      legend.background  = element_rect(fill = "#0b1520", color = NA),
      legend.key         = element_rect(fill = "#0b1520", color = NA),

      # Axes
      axis.line          = element_line(color = "#1a2e42", linewidth = 0.4),
      axis.ticks         = element_line(color = "#1a2e42", linewidth = 0.3),

      # Margins
      plot.margin        = margin(20, 20, 20, 20)
    )
}

# Set as default theme
theme_set(theme_cytoCrunch())

# CytoCrunch accent palette for plots
cc_palette <- c(
  "#00d4ff", "#7b2fff", "#00ff9d", "#ff6b6b",
  "#ffd93d", "#ff922b", "#4d96ff", "#6bcb77",
  "#e066ff", "#00b8a9", "#f8a5c2", "#78e08f",
  "#38ada9", "#f6b93b", "#e55039", "#4a69bd"
)
```

---

# Overview

**CytoCrunch** is an R package that bridges FlowJo's plugin ecosystem and Bioconductor workflows for flow cytometry data analysis. It provides a unified S4 object system for managing, transforming, clustering, and visualizing high-parameter cytometry data at scale.

This notebook demonstrates the core analytical pipeline:

1. **Data loading** — importing FlowJo-exported CSV data into CytoCrunch objects
2. **Clustering** — FlowSOM-based automated cell population discovery
3. **Dimensionality reduction** — t-SNE and UMAP embeddings for visualization
4. **Visualization** — publication-ready plots including embeddings, heatmaps, and merged multi-sample views

<div class="callout-note">
<strong>Note</strong> — All data shown uses the built-in demo dataset: 3,000 cells each from four tissue types (Blood, Liver, Lymph Node, Spleen), exported from FlowJo as extended node CSVs.
</div>

---

# Setup

## Load CytoCrunch and Dependencies

```{r load-packages}
devtools::load_all(here::here())
library(ggplot2)
library(gridExtra)
```

---

# Data Loading

## The CytoCrunch S4 Object

CytoCrunch uses a custom **S4 object system** to encapsulate all analysis state — raw data, clustering results, embeddings, and plots — in a single, self-contained object. This design ensures reproducibility and enables seamless chaining of analytical steps.

Each `CytoCrunch` object is initialized from a FlowJo-exported CSV file:

```{r load-data}
# ── Load all four demo datasets in one call ───────────────────────────────────
demo      <- loadDemoData()
blood     <- demo$blood
lymphNode <- demo$lymphNode
liver     <- demo$liver
spleen    <- demo$spleen
```

## Inspect the Object

```{r inspect-object}
# Object structure — S4 slots
slotNames(blood)

# Sample name
blood@sampleName

# Preview the expression matrix dimensions
dim(blood@inputCSV$X)
```

```{r preview-data}
# Preview the first few rows and marker columns
head(blood@inputCSV$X[, 1:8])
```

## Available Markers

```{r list-markers}
# Derive display labels: stain name where available, else parameter name
stainNames   <- blood@parameterNames$fjStainNames
parNames     <- blood@parameterNames$fjParNames
markerLabels <- ifelse(is.na(stainNames), parNames, stainNames)
markerLabels
```

---

# Clustering

## FlowSOM Clustering

CytoCrunch uses **FlowSOM** for automated, unsupervised cell population discovery. The `performClustering()` function wraps the FlowSOM algorithm and stores results directly in the S4 object, tagged with a unique `runID` for tracking multiple analyses.

### Blood — 12 Clusters

```{r cluster-blood}
blood <- performClustering(blood,
                           runID  = "BLOOD_C12",
                           method = "FlowSOM",
                           k      = 12)
```

### Lymph Node — 15 Clusters

```{r cluster-lymphnode}
lymphNode <- performClustering(lymphNode,
                               runID  = "LN_C15",
                               method = "FlowSOM",
                               k      = 15)
```

### Liver — 12 Clusters

```{r cluster-liver}
liver <- performClustering(liver,
                           runID  = "LIVER_C12",
                           method = "FlowSOM",
                           k      = 12)
```

### Spleen — 12 Clusters

```{r cluster-spleen}
spleen <- performClustering(spleen,
                            runID  = "SPLEEN_C12",
                            method = "FlowSOM",
                            k      = 12)
```

<div class="callout-note">
<strong>Run IDs</strong> — Each clustering and embedding call is tagged with a `runID` string. This lets you store multiple analyses (different `k` values, methods, parameters) on the same object and reference them independently downstream.
</div>

---

# Dimensionality Reduction

## Embedding Methods

CytoCrunch supports both **t-SNE** and **UMAP** for 2D embedding of high-dimensional cytometry data. Like clustering, embeddings are stored on the object and referenced by `runID`.

### Blood — t-SNE

```{r embed-blood-tsne}
blood <- performEmbedding(blood,
                          runID    = "BLOOD_C12",
                          method   = "tsne",
                          fjSeed   = 123,
                          fjXdim   = 10,
                          fjYdim   = 10,
                          fjRLEN   = 15,
                          fjK      = 10,
                          fjAdjust = 0,
                          fjSmooth = 0,
                          fjMapSize = 2000)
```

### Lymph Node — UMAP

```{r embed-lymphnode-umap}
lymphNode <- performEmbedding(lymphNode,
                              runID      = "LN_C15",
                              method     = "umap",
                              fjSeed     = 42,
                              fjXdim     = 10,
                              fjYdim     = 10,
                              fjRLEN     = 22,
                              fjK        = 12,
                              fjAdjust   = 0,
                              fjSmooth   = 1,
                              fjUmap.Dist = 0.4,
                              fjMapSize   = 2000)
```

### Liver — UMAP

```{r embed-liver-umap}
liver <- performEmbedding(liver,
                          runID      = "LIVER_C12",
                          method     = "umap",
                          fjSeed     = 77,
                          fjXdim     = 10,
                          fjYdim     = 10,
                          fjRLEN     = 20,
                          fjK        = 10,
                          fjAdjust   = 0,
                          fjSmooth   = 1,
                          fjUmap.Dist = 0.3,
                          fjMapSize   = 2000)
```

### Spleen — UMAP

```{r embed-spleen-umap}
spleen <- performEmbedding(spleen,
                           runID      = "SPLEEN_C12",
                           method     = "umap",
                           fjSeed     = 99,
                           fjXdim     = 10,
                           fjYdim     = 10,
                           fjRLEN     = 18,
                           fjK        = 11,
                           fjAdjust   = 0,
                           fjSmooth   = 1,
                           fjUmap.Dist = 0.35,
                           fjMapSize   = 2000)
```

---

# Visualization

## Embedding Plots with Cluster Overlay

The `plotEmbeddingWithClusters()` function renders 2D embeddings colored by cluster assignment — the core visualization for evaluating how well FlowSOM has resolved distinct cell populations.

### Blood (t-SNE)

```{r plot-blood-embed, fig.height=7}
plotEmbeddingWithClusters(blood,
                          embeddingRunID  = "BLOOD_C12",
                          clusteringRunID = "BLOOD_C12")
```

### Lymph Node (UMAP)

```{r plot-ln-embed, fig.height=7}
plotEmbeddingWithClusters(lymphNode,
                          embeddingRunID  = "LN_C15",
                          clusteringRunID = "LN_C15")
```

### Liver (UMAP)

```{r plot-liver-embed, fig.height=7}
plotEmbeddingWithClusters(liver,
                          embeddingRunID  = "LIVER_C12",
                          clusteringRunID = "LIVER_C12")
```

### Spleen (UMAP)

```{r plot-spleen-embed, fig.height=7}
plotEmbeddingWithClusters(spleen,
                          embeddingRunID  = "SPLEEN_C12",
                          clusteringRunID = "SPLEEN_C12")
```

---

## Cluster Heatmaps

The `plotClusterHeatmap()` function generates marker-by-cluster heatmaps — critical for **phenotypic interpretation** of discovered populations. By examining which markers are high or low in each cluster, you can assign biological identity (e.g., CD3+CD8+ → cytotoxic T cells).

### Full Marker Panel (Row-Scaled)

```{r heatmap-full}
blood <- plotClusterHeatmap(blood, clusteringRunID = "BLOOD_C12")

# Verify storage in the S4 plot list
cat("Plots stored:", length(blood@plotList), "\n")
cat("Plot type:   ", blood@plotList[[1]]@plotType, "\n")
cat("Run ID:      ", blood@plotList[[1]]@runID, "\n")
```

### Selected Immune Markers (Unscaled)

```{r heatmap-subset}
blood <- plotClusterHeatmap(blood,
                            clusteringRunID = "BLOOD_C12",
                            markers = c("CD45", "HLA-DR", "CD16", "CD8a",
                                        "CD4", "CD3", "CD19", "CD56",
                                        "CD127", "CD95", "NKG2D"),
                            scale = "none")

cat("Total plots stored:", length(blood@plotList), "\n")
```

---

## Merged Multi-Tissue Analysis

One of CytoCrunch's key capabilities is **merging multiple CytoCrunch objects** for cross-tissue comparison. The `mergeCytoCrunchObjects()` function concatenates expression data, embeddings, and cluster assignments while preserving sample provenance.

```{r merge-objects}
# Merge Blood and Lymph Node for cross-tissue comparison
mergedCC <- mergeCytoCrunchObjects(
  list(blood, lymphNode),
  c("BLOOD_C12", "LN_C15")
)
```

```{r plot-merged, fig.height=7}
plotMergedEmbeddings(mergedCC)
```

<div class="callout-note">
<strong>Cross-Tissue Insight</strong> — The merged embedding reveals how immune cell populations from different tissue compartments occupy shared or distinct regions of phenotypic space. Overlapping clusters may indicate circulating populations, while tissue-exclusive clusters suggest resident or specialized cell types.
</div>

---

# Session Info

```{r session-info}
sessionInfo()
```

---

<div style="text-align:center; margin-top:4rem; padding:2rem 0; border-top:1px solid #1a2e42;">
<p style="font-family:'Space Mono', monospace; font-size:0.75rem; letter-spacing:0.12em; text-transform:uppercase; color:#3a5a70;">
CytoCrunch · Jason Luthy · <a href="https://github.com/jluthy/CytoCrunch" style="color:#00d4ff !important;">github.com/jluthy/CytoCrunch</a>
</p>
</div>
